---
title: 'Reproducible Research: Exploring Storm Data impacts on people and
  economics'
author: "Jeanna"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_notebook: default
  word_document: default
  html_document:
    df_print: paged
fontsize: 11pt
urlcolor: blue
linkcolor: red
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tinytex)
# tinytex::install_tinytex()
# to uninstall TinyTeX, run
# tinytex::uninstall_tinytex()

```

**Synopsis**

Storm and other severe weather events can cause both public health and economic problems for communities and municipalities.  
In this project, we explore [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) from the U.S. National Oceanic and Atmospheric Administration(NOAA). This contains storms and weather events in the United States with time, location of the occurrences, estimations of any fatalities, injuries, and property damage. We will analyze and find out what type of events have big impacts for population health and economic consequences. 

## Preparation{#prep}

We created two mapping csv files manually and store them in folder /data under project folder. They are used to clean event type data. For mapping data detail, they are listed in the [Appendix](#apdx).  
- **ref_Evtype_Corrections.csv** is for correcting the typos of original event type  
- **ref_Evtype_Mapping.csv** is for mapping the official events type  

Before data processing, load R packages.

```{r package_options, warning=FALSE, message=FALSE}

## Loading packages
library(dplyr)
library(ggplot2)
library(lubridate)
library(stringr)
library(tidyr)
library(cowplot)

```


## Data Processing  
In this section, Data will be processed by five steps.

- Loading data
- Converting date character to date type and reducing the data size by aggregating
- Cleaning Event Type
- Converting property damage and crop damage to million
- Calculating data

### Loading data  
  
```{r, warning=FALSE, message=FALSE }
## To denote the starting time
startTime <- Sys.time()

filename <- "data/repdata_data_StormData.csv.bz2"
if (!file.exists(filename)) {
    file_url <-
        "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
    download.file(file_url, filename, method = "curl")
}
## Reading storm data set and processing the data
df.storm <- read.csv(filename)
cat("In the data file, there are",
    nrow(df.storm),
    "records on",
    ncol(df.storm),
    "variables",
    "\n"
)
```

### Converting date character to date type and reducing the data size by aggregating  

Convert date character to date type
 
```{r, warning=FALSE, message=FALSE }
## Convert date character to date type
df.storm$BGN_DATE <- as.Date(df.storm$BGN_DATE, format = "%m/%d/%Y")
df.storm$END_DATE <- as.Date(df.storm$END_DATE, format = "%m/%d/%Y")
## Add Year column
df.storm$BGN_YEAR <- as.character(df.storm$BGN_DATE, format = "%Y")
```

To reduce data, aggregate fatalities, injuries, crop damage and property damage by PROPDMGEXP, CROPDMGEXP, event type and year.

```{r, warning=FALSE, message=FALSE }

## Aggregate fatalities, injuries, and property damage 
## by event type and beginning year
df.byevtypeyear <-
    df.storm %>% group_by(EVTYPE, BGN_YEAR, PROPDMGEXP, CROPDMGEXP) %>%
    summarize(
        sumFatalities = sum(FATALITIES),
        sumInjuries = sum(INJURIES),
        sumPropdmg = sum(PROPDMG),
        sumCropdmg = sum(CROPDMG)
    )

cat("after reduced, the data rows is",
    nrow(df.byevtypeyear),
    "\n"
)
checkDataType <- unique(df.byevtypeyear[, "EVTYPE"])
cat ("The official events type are 48.",
    "However, 'EVTYPE' column has", nrow(checkDataType), "unique events",
    "\n"
)

```

### Cleaning Event Type

TO simplify data cleaning process, two configuration files are created manually (mentioned in Section [Preparation](#prep)). Firstly, read file ref_Evtype_Corrections.csv and correct the typos of event type. Then, after loading data from file ref_Evtype_Mapping.csv, map event type to the official events type and remove the data without official events type.
  
```{r, warning=FALSE, message=FALSE }

## Convert EVTYPE to only the first letter to Uppercase and remove whitespace
dfNew <- df.byevtypeyear
dfNew$EVTYPE <- str_trim(str_to_title(dfNew$EVTYPE))
## convert backslash \ to forward slash / 
dfNew$EVTYPE <- gsub("\\\\", "/", dfNew$EVTYPE )

## Loading Typo Mapping data manually created for correcting typos
refTypo <- read.csv("data/ref_Evtype_Corrections.csv", header = TRUE)
for (i in 1:nrow(refTypo)) {
    dfNew$EVTYPE <-
        sub(refTypo$EVTXT[i],
            refTypo$EVSUBSTR[i],
            dfNew$EVTYPE)
}

## Loading Event Type Mapping data manually created,
## which are used for simplifying remove operation and grepl operation
refEvType <- read.csv("data/ref_Evtype_Mapping.csv", header = TRUE)
## Retrieving subset event type data for removing operation and processing data
refEvTypeRemove <- refEvType [refEvType$OPER=="remove", ]
for (i in 1:nrow(refEvTypeRemove)) {
    patx <- refEvTypeRemove$SEARCHEVTYPE [i]
    searchLoc <- grepl(patx, dfNew$EVTYPE)
    dfNew <- dfNew[!(searchLoc), ]
}
## Fetching and sorting Event Type for grepl operation by priority
refEvTypeGrepl <- refEvType [refEvType$OPER=="grepl", ]
refEvTypeGrepl <- arrange(refEvTypeGrepl, PRIO, STORMEVENT)
## updating storm event with official one.
dfNew["StormEvent"] <- NA
for (i in 1:nrow(refEvTypeGrepl)) {
    patx <- refEvTypeGrepl$SEARCHEVTYPE [i]
    searchLoc <- grepl(patx, dfNew$EVTYPE)
    dfNew$StormEvent[is.na(dfNew$StormEvent) &
                       searchLoc == TRUE] <- refEvTypeGrepl$STORMEVENT[i]
}
## Remove rows with NA StormEvent
dfNew <- dfNew[!(is.na(dfNew$StormEvent)), ]

cat(
    "Removed",
    nrow(df.byevtypeyear) - nrow(dfNew),
    "uncount records, and left data has",
    nrow(dfNew),
    "records.",
    "\n"
)

checkDataType <- unique(dfNew[ , c("StormEvent")]) 
cat ("Currently, the storm events type are", 
     nrow(checkDataType), 
     "\n"
)

```

### Converting property damage and crop damage to million  
For the value of column 'CROPDMGEXP' and 'PROPDMGEXP', we refer to a nice work [How To Handle Exponent Value of PROPDMGEXP and CROPDMGEXP](https://rstudio-pubs-static.s3.amazonaws.com/58957_37b6723ee52b455990e149edde45e5b6.html) which mentioned in discussion forum by mentor Usama.

- H,h = hundreds = 100
- K,k = kilos = thousands = 1,000
- M,m = millions = 1,000,000
- B,b = billions = 1,000,000,000
- (+) = 1
- (-) = 0
- (?) = 0
- black/empty character = 0
- numeric 0..8 = 10
 
```{r, warning=FALSE, message=FALSE }
dfNew$PROPDMGEXP <- toupper(dfNew$PROPDMGEXP)
dfNew$CROPDMGEXP <- toupper(dfNew$CROPDMGEXP)

iNum <- 0:8
dfNew <- mutate(
    dfNew,
    PROPDMG_M = case_when(
        PROPDMGEXP == "B" ~  sumPropdmg * 1000,
        PROPDMGEXP == "M" ~  sumPropdmg,
        PROPDMGEXP == "K" ~  sumPropdmg / 1000,
        PROPDMGEXP == "H" ~  sumPropdmg / 10000,
        PROPDMGEXP %in% iNum ~ sumPropdmg / 100000,
        PROPDMGEXP == "+" ~ sumPropdmg / 1000000,
        TRUE ~ 0
    )
)
dfNew <- mutate(
    dfNew,
    CROPDMG_M = case_when(
        CROPDMGEXP == "B" ~  sumCropdmg * 1000,
        CROPDMGEXP == "M" ~  sumCropdmg,
        CROPDMGEXP == "K" ~  sumCropdmg / 1000,
        CROPDMGEXP == "H" ~  sumCropdmg / 10000,
        CROPDMGEXP %in% iNum ~ sumCropdmg / 100000,
        CROPDMGEXP == "+" ~ sumCropdmg / 1000000,
        TRUE ~ 0
    )
)

```

### Calculating data  

Summarize the data by official event type.
  
```{r, warning=FALSE, message=FALSE }
byEvtype <- dfNew %>%
    group_by(StormEvent) %>%
    summarise(
        DMG = round(sum(PROPDMG_M) + sum(CROPDMG_M)),
        FATALITIES = sum(sumFatalities),
        INJURIES = sum(sumInjuries)
    )
byEvtypeYear <- dfNew %>%
    group_by(StormEvent, BGN_YEAR) %>%
    summarise(
        DMG = round(sum(PROPDMG_M) + sum(CROPDMG_M)),
        FATALITIES = sum(sumFatalities),
        INJURIES = sum(sumInjuries)
    )
## Pivot data from wide to long
byEvTypeEcon <- top_n(byEvtype, 10, DMG)
byEvTypeHlth <- top_n(byEvtype, 10, (FATALITIES + INJURIES))
byEvtypeHlthPiv <-
    subset(byEvTypeHlth, select = -DMG) %>% 
      pivot_longer(-StormEvent,
                   names_to = "TYPE",
                   values_to = "CONSEQVALUE")


```

## Results  

Plot the top 10 injuries and fatalities, and top 10 sum of crop damage property damage to check which type of event have the most impact.    

```{r, warning=FALSE, message=FALSE}
g <-
    ggplot(byEvtypeHlthPiv, aes(StormEvent, CONSEQVALUE, fill = TYPE)) +
    geom_bar(stat = "identity") +
    theme(axis.text.x = element_text(angle = 90),
          plot.title = element_text(hjust = 0.5),
          legend.title = element_blank(),
          legend.position = "bottom" ) +
    ggtitle("The Top 10 Events Harmful to Population Health") +
    labs(x = "Storm Data Event", y = "Impacted Number of Individuals")
print(g)
```

From above diagram, we can conclude that Tornado is the most harmful storm event to population health. 

```{r, warning=FALSE, message=FALSE}
## Histogram the top 10 harmful events to economic consequences
g <- ggplot(data = byEvTypeEcon, aes(x = StormEvent, y = DMG, fill = DMG)) +
    geom_bar(stat = "identity") +
    theme(axis.text.x = element_text(angle =  70, vjust = 0.5),
          plot.title = element_text(hjust = 0.5)) +
    scale_fill_gradient(low="pink",high="navy") +
    ggtitle("The Top 10 Events to Economic Damage") +
    labs(x = "Storm Data Event", y = "Damage (Million)")
print(g)

```

From above diagram, we can conclude that Flood have the greatest economic consequences.

Additionally, we would like to check what is the impacts over years for the first two most influential events, for people, the first two events are Tornado/Heat, and for economics, they are Flood/Hurricane (Typhoon). 

```{r, warning=FALSE, message=FALSE}
## Check Storm Event Tornado/Flood/Heat/Hurricane (Typhoon) impacts
byEvtypeYearTop <-
    subset(
        byEvtypeYear,
        StormEvent == "Tornado" |
            StormEvent == "Flood" |
            StormEvent == "Hurricane (Typhoon)" |
            StormEvent == "Heat"
    )

plot.yHlth <-
    ggplot(byEvtypeYearTop,
           aes(x = BGN_YEAR,
               y = FATALITIES + INJURIES,
               group = StormEvent
           )) +
    geom_line(aes(color = StormEvent), size = 1)+
    geom_point(aes(color = StormEvent), size = 1.5)+
    scale_x_discrete(breaks = seq(1950, 2015, by = 5)) +
    theme(
        axis.text.x = element_text(angle = 70, vjust = 0.5),
        legend.title = element_blank(),
        legend.position = "bottom"
    ) +
    labs(x = "Year", y = "Fatalities & Injuries")

plot.yEcon <-
    ggplot(data = byEvtypeYearTop,
           aes(x = BGN_YEAR, y = DMG, fill = StormEvent)) +
    # geom_point(size = 3, alpha = 0.5) +
    geom_bar(stat = "identity") +
    theme(
        axis.text.x = element_text(angle = 70, vjust = 0.5),
        legend.title = element_blank(),
        legend.position = "bottom"
    ) +
    scale_x_discrete(breaks = seq(1950, 2015, by = 5)) +
    labs(x = "Year", y = "Damage (Million)")

gg <-
    ggdraw() + 
    draw_plot(plot.yHlth, 0, 0, 0.5, 1) + 
    draw_plot(plot.yEcon, 0.5, 0, 0.5, 1) 
print(gg)



```

From the plots, we can see Tornado always has impacts on people and economics since 1950. Since 1993, Flood and began to affect on economics and population's health.  

```{r, warning=FALSE, message=FALSE}
## To denote the end time
endTime <- Sys.time()
cat("Processing time:", endTime-startTime, "\n")
```

\newpage

## Appendix{#apdx}  
File **ref_Evtype_Corrections.csv** is for correcting the typos of original event type.
  

```{r tables-refTypo, warning=FALSE, message=FALSE, echo=FALSE}
knitr::kable(refTypo, caption = 'Mapping the Typos Correction')
```

File **ref_Evtype_Mapping.csv** is for mapping official events type.

```{r tables-refEvType, warning=FALSE, message=FALSE, echo=FALSE}
knitr::kable(refEvType, caption = 'Mapping the official events type') 

```
